---
title: "Project ALS — Natural Language Access to the Variant Database"
subtitle: "From a single-table prototype to a six-server agentic pipeline"
author: "Luuk Engels, Robin Jansen, Siard Groot — Projecticum Data Science"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
    toc_depth: 3
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 6
)

library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(knitr)
library(kableExtra)
library(DiagrammeR)

# ── Path to the final graded benchmark CSV ──────────────────
GRADED_CSV <- "~/project_ALS_databse/analysis/benchmark_surf_grading/grading_20260621_143922/finalgraded_20260621_121012.csv"
# ──────────────────────────────────────────────────────────────

combined <- read.csv(GRADED_CSV)
combined$grade_answer <- ifelse(toupper(as.character(combined$grade_answer)) == "TRUE", 1L,
                          ifelse(toupper(as.character(combined$grade_answer)) == "FALSE", 0L,
                          suppressWarnings(as.integer(combined$grade_answer))))

ALL_CATEGORIES <- c("simple","analytical","complex","phenotype","annotation_trap",
                     "unanswerable","rvat","nonsense","toolfree")
ALL_CATEGORY_LABELS <- c("Simple (S)","Analytical (A)","Complex (C)","Phenotype (P)",
                          "Annotation Trap (T)","Unanswerable (U)","RVAT (R)",
                          "Nonsense (N)","Toolfree (F)")

theme_als <- theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 15, margin = margin(b = 10)),
    plot.subtitle    = element_text(color = "grey40", size = 11, margin = margin(b = 10)),
    axis.title       = element_text(face = "bold"),
    axis.text        = element_text(color = "grey20"),
    axis.text.x      = element_text(angle = 45, hjust = 1),
    legend.title     = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey90"),
    plot.caption     = element_text(color = "grey50", size = 9, margin = margin(t = 10))
  )
```

> **This document is the ground truth for the live presentation.** Everything
> shown live on the SURF server — the working app, example questions, the
> architecture walkthrough — is documented here in full, with the actual
> benchmark numbers and reasoning behind each design decision. Find this file,
> and everything it references, in the project's git repository.

---

# 0. The Goal, in One Sentence

Let a researcher ask the Project MinE ALS genomic variant database a question in **plain English** — no SQL, no knowledge of the `rvat` R package — and get back a **correct, grounded** answer.

Everything in this document is the story of how far that goal actually got pushed, told through four distinct stages of the project.

---

# Stage 1 — A Single Table, One Model, One Tool

## Where it started

The very first working prototype answered questions against **exactly one table**: `varInfo_synthetic`, a synthetic 10-sample genotype subset (5 ALS cases, 5 matched controls) covering 1,802 variants across 12 ALS-associated genes. One Ollama model. One query path. No routing, no multi-step reasoning, no fallback if the question needed anything beyond a single SQL query against that one table.

```{r stage1-diagram, echo=FALSE, fig.cap="Stage 1 — single table, single tool"}
grViz("
digraph stage1 {
  graph [layout = dot, rankdir = LR, fontname = 'Helvetica', fontsize = 12, bgcolor = 'white']
  node  [fontname = 'Helvetica', fontsize = 12, style = filled, shape = roundedbox]

  User    [label = 'Researcher\\n(plain English)', fillcolor = '#E8F4FD', color = '#457B9D']
  LLM     [label = 'Local LLM\\n(Ollama)',         fillcolor = '#E1D5E7', color = '#9B5DE5']
  QC      [label = 'querychat\\n(NL to SQL)',       fillcolor = '#FFF2CC', color = '#E9C46A']
  DB      [label = 'varInfo_synthetic\\n(one table)', fillcolor = '#FFE6CC', color = '#F4A261']
  Answer  [label = 'Answer',                       fillcolor = '#D5E8D4', color = '#2A9D8F']

  User -> LLM -> QC -> DB -> Answer
}
")
```

## What this version could and could not do

This worked, in a narrow sense — for simple questions wholly contained within that one table, it produced reasonable answers. But the limitation was structural, not a bug to be patched: there was no way to ask anything that required phenotype data (sex, age, population), no way to run a real statistical test, and no graceful way to refuse a question the table genuinely couldn't answer. A single querychat connection is fundamentally a single-table tool.

This is exactly where the project's first major pivot came from.

---

# Stage 2 — Reaching Across Tables

## The problem

A huge fraction of genuinely interesting ALS research questions need **more than one table at once** — "how many female carriers of a given gene are there," for instance, requires joining the genotype table to a separate phenotype table containing sex, age, and population data for the full cohort.

`querychat` as a single-source tool couldn't do this natively. The team explored two paths to close this gap: a custom `MultiTableSource` extension to `querychat` itself, and, in parallel, a direct exploration of `ellmer` with hand-defined tools — which turned out to be the more flexible foundation for what came next.

## What this stage actually unlocked

By the end of this stage, the prototype could correctly join genotype columns to the **`pheno`** table — a much larger, 25,000-sample real cohort, not just the small synthetic genotype subset — and answer questions involving sex, population, and age, alongside the variant-level data from Stage 1.

This is also where the team confirmed something important: the small "IID" naming convention linking genotype columns (`ALS_1`, `Control_3`, …) to phenotype rows required deliberate, explicit handling — `ALS_1` in the genotype table corresponds to IID `'ALS1'` (no underscore) in the phenotype table. Small detail, but it was a real source of early bugs, and it's documented explicitly in the final tool code today specifically because of how much friction it caused here.

---

# Stage 3 — Going Agentic: Multiple Servers, Real Tool-Calling

## Why a bigger architectural change was needed

Multi-table access solved one limitation. It didn't solve two others: the system still had no way to **decide which of several different tool types** a question actually needed (a plain SQL lookup? A phenotype join? A real statistical test?), and it still had no mechanism to **gracefully refuse** a genuinely unanswerable question rather than guessing.

This is where the project moved from a single querychat pipeline to a proper **agentic, multi-server architecture** using the Model Context Protocol (MCP) — giving the model a defined set of callable tools, spread across multiple specialized servers, plus the judgment to pick between them.

```{r stage3-diagram, echo=FALSE, fig.cap="Stage 3 — agentic architecture, multiple MCP servers"}
grViz("
digraph stage3 {
  graph [layout = dot, rankdir = TB, fontname = 'Helvetica', fontsize = 11, bgcolor = 'white', ranksep = 0.4]
  node  [fontname = 'Helvetica', fontsize = 11, style = 'filled,rounded', shape = box]
  edge  [fontname = 'Helvetica', fontsize = 9, color = '#888888', fontcolor = '#555555']

  researcher [label = 'Researcher', fillcolor = '#C6E8DF', color = '#0A5C47', fontcolor = '#0A5C47']
  llm        [label = 'LLM orchestrator\\n(picks a tool)', fillcolor = '#ECEAFB', color = '#4A42B0', fontcolor = '#302880']
  mcpo       [label = 'mcpo (MCP to HTTP proxy)', fillcolor = '#FDF0DC', color = '#9A6010', fontcolor = '#6B4200']
  variant    [label = 'variant_analysis', fillcolor = '#FDF0DC', color = '#9A6010', fontcolor = '#6B4200']
  geno       [label = 'genotype_analysis', fillcolor = '#FDF0DC', color = '#9A6010', fontcolor = '#6B4200']
  pheno      [label = 'phenotype_data', fillcolor = '#FDF0DC', color = '#9A6010', fontcolor = '#6B4200']
  clinvar    [label = 'clinvar_annotation', fillcolor = '#FDF0DC', color = '#9A6010', fontcolor = '#6B4200']
  db         [label = 'rvatData.gdb', shape = cylinder, fillcolor = '#ECEAE3', color = '#7A7870', fontcolor = '#3C3C3A']
  answer     [label = 'Answer', fillcolor = '#C6E8DF', color = '#0A5C47', fontcolor = '#0A5C47']

  researcher -> llm -> mcpo
  mcpo -> variant -> db
  mcpo -> geno -> db
  mcpo -> pheno -> db
  mcpo -> clinvar -> db
  db -> mcpo -> answer
}
")
```

## What this unlocked

By this point the system could: route a question to exactly the right tool out of several specialized servers; refuse an unanswerable question explicitly rather than hallucinating; and — critically — generate **its own SQL** for the long tail of questions no pre-built tool covered, with that SQL validated and, if it failed, **self-corrected** using the real database error fed back to the model.

This is also where the project's testing rigor stepped up: a benchmark of dozens of questions grew into the full **111-question, 9-category benchmark** still in use today, because a multi-tool architecture has many more ways to go subtly wrong than a single querychat call, and informal testing stopped being enough to trust the system.

---

# Stage 4 — Real Statistics: the `rvat` Bridge

## The last major capability gap

Even with multiple tools and self-correcting SQL, one whole category of genuinely important questions remained out of reach: **real statistical genetics**. "Does this gene show a statistically significant burden of rare variants in ALS cases versus controls?" is not a SQL question — it requires the actual `rvat` R package's burden-testing machinery (Firth regression, SKAT, minor allele frequency calculated properly from a genotype matrix, linkage disequilibrium).

## The fix: bridging into a real R statistical server

The final architecture adds a dedicated **R/plumber HTTP server** running `rvat` directly, with a thin Python MCP bridge (`rvat_bridge.py`) exposing its endpoints as callable tools alongside everything else. This is the piece that turns the system from "a natural-language SQL interface" into something that can genuinely support real genomic research questions: burden tests, MAF, LD, single-variant association — computed by the actual statistical package researchers would otherwise run by hand.

```{r stage4-diagram, echo=FALSE, fig.cap="Stage 4 — final architecture, six MCP servers including the rvat bridge"}
grViz("
digraph stage4 {
  graph [layout = dot, rankdir = TB, fontname = 'Helvetica', fontsize = 11,
         ranksep = 0.4, nodesep = 0.45, bgcolor = 'white']
  node  [fontname = 'Helvetica', fontsize = 11, style = 'filled,rounded',
         shape = box, penwidth = 1.1, margin = '0.18,0.1']
  edge  [fontname = 'Helvetica', fontsize = 9, penwidth = 1.0, color = '#888888', fontcolor = '#555555']

  researcher [label = 'Researcher', fillcolor = '#C6E8DF', color = '#0A5C47', fontcolor = '#0A5C47']
  shiny      [label = 'final_app.R\\n(Shiny frontend)', fillcolor = '#C6E8DF', color = '#0A5C47', fontcolor = '#0A5C47']
  gate       [label = 'Nonsense / unanswerable gate\\n(no LLM call if caught)', fillcolor = '#FBE3E3', color = '#A23B3B', fontcolor = '#7A2424']
  classify   [label = 'classify_complexity()\\nsimple vs complex', fillcolor = '#ECEAFB', color = '#4A42B0', fontcolor = '#302880']
  decompose  [label = 'decompose_question()\\n(skipped when not needed)', fillcolor = '#ECEAFB', color = '#4A42B0', fontcolor = '#302880', style = 'filled,rounded,dashed']
  route      [label = 'route_question()\\npicks server + tool', fillcolor = '#ECEAFB', color = '#4A42B0', fontcolor = '#302880']
  sqlgen     [label = 'generate_sql() + validate_sql()', fillcolor = '#F5F3FF', color = '#7060C8', fontcolor = '#302880', style = 'filled,rounded,dashed']
  agentic    [label = 'run_agentic_pipeline()\\nmulti-step, max 5 steps', fillcolor = '#FFF3CD', color = '#8A6D1A', fontcolor = '#5C4A10', style = 'filled,rounded,dashed']
  mcpo       [label = 'mcpo (MCP to HTTP proxy)', fillcolor = '#FDF0DC', color = '#9A6010', fontcolor = '#6B4200']
  tools      [label = '6 MCP servers', fillcolor = '#FDF0DC', color = '#9A6010', fontcolor = '#6B4200']
  rvat       [label = 'rvat plumber server (R)\\nburden tests, MAF, LD', fillcolor = '#E3EEFB', color = '#2A5C9A', fontcolor = '#1A3C6B']
  db         [label = 'rvatData.gdb\\nvarInfo_synthetic + pheno', shape = cylinder, fillcolor = '#ECEAE3', color = '#7A7870', fontcolor = '#3C3C3A']
  selfcorrect[label = 'Self-correct on SQL error\\n(real error fed back, 1 retry)', fillcolor = '#FBE3E3', color = '#A23B3B', fontcolor = '#7A2424', style = 'filled,rounded,dashed']
  summarize  [label = 'summarize_result()\\n1-sentence answer', fillcolor = '#C6E8DF', color = '#0A5C47', fontcolor = '#0A5C47']
  answer     [label = 'Answer + data table', fillcolor = '#C6E8DF', color = '#0A5C47', fontcolor = '#0A5C47']

  researcher -> shiny -> gate
  gate -> classify [label = 'passes']
  gate -> answer [label = 'caught - instant reply', style = dashed]
  classify -> decompose [label = 'simple']
  classify -> agentic [label = 'complex']
  decompose -> route -> sqlgen [label = 'if SQL tool', style = dashed]
  route -> mcpo
  sqlgen -> mcpo
  agentic -> mcpo
  mcpo -> tools -> db
  tools -> rvat [style = dashed]
  rvat -> db
  tools -> selfcorrect [label = 'on error', style = dashed]
  selfcorrect -> tools [label = 'retry', style = dashed]
  tools -> summarize -> answer
}
")
```

## A simplification that happened along the way

One detail worth being upfront about: an earlier version of this final stage ran **two models** — a large orchestrator plus a smaller, dedicated SQL-writing model. Careful benchmarking, plus a direct code review, showed the second model wasn't reliably improving results, and in some configurations wasn't even properly wired into the decision logic. The architecture was simplified to a **single model, llama3.1:70b, throughout** — and this simpler version benchmarked *better*, not worse, partly because the SURF server has no GPU, so two models competing for the same CPU cores was actively working against the system rather than helping it.

---

# The Benchmark: How We Know Any of This Actually Works

```{r tool-table, echo=FALSE}
tool_table <- data.frame(
  Server = c("db_exploration", "variant_analysis", "genotype_analysis",
             "phenotype_data", "clinvar_annotation", "rvat_analysis"),
  Covers = c(
    "Schema, table info, explicit database limitations",
    "Variant-level SQL: counts, impact filters, CADD/SIFT/PolyPhen, allele frequency, gene summaries",
    "Carrier and burden analysis: per-sample burden, threshold-based carrier counts, case/control dosage ratios",
    "Multi-table joins between genotype columns and the pheno table",
    "ClinVar lookup via rsID, with a local cache",
    "Bridge to the R/plumber rvat server — real burden tests, MAF, LD, association tests, full-cohort carrier counts"
  )
)
kable(tool_table, caption = "The six MCP servers in the final architecture") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```

```{r category-table, echo=FALSE}
cat_table <- data.frame(
  Category = c("Simple (S)", "Analytical (A)", "Complex (C)", "Phenotype (P)",
               "Annotation Trap (T)", "Unanswerable (U)", "RVAT (R)",
               "Nonsense (N)", "Toolfree (F)"),
  Count = c(16, 17, 18, 6, 3, 15, 5, 16, 15),
  Tests = c(
    "Single-fact lookups",
    "Aggregation, ratios, comparisons",
    "Multi-step reasoning, per-entity breakdowns",
    "Joins between genotype and phenotype tables",
    "Using the actual coded value, not its English description",
    "Correctly refusing a genuine data gap, without hallucinating",
    "Real statistical tests via the rvat R package",
    "Refusing impossible questions, asking for clarification on vague ones",
    "Complex SQL answerable without specialised tools"
  )
)
kable(cat_table, caption = "111 benchmark questions across 9 categories") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```

Every response is graded by an LLM grader against a written ground-truth answer key, on four independent criteria: **answer correctness**, **conciseness**, **hallucination-free**, and **tool correctness**.

---

# Final Results

```{r overall-score, echo=FALSE, fig.height=3}
overall <- mean(combined$grade_total, na.rm = TRUE)
ggplot(data.frame(model = "llama3.1:70b [adaptive]", score = overall),
       aes(x = model, y = score)) +
  geom_col(fill = "#2A9D8F", width = 0.5) +
  geom_text(aes(label = sprintf("%.2f", score)), hjust = -0.3, size = 6, fontface = "bold") +
  coord_flip() +
  scale_y_continuous(limits = c(0, 4.3), breaks = 0:4) +
  labs(title = "Overall average score", subtitle = "Averaged across all 111 benchmark questions",
       x = NULL, y = "Average score (out of 4)") +
  theme_als
```

**3.51 / 4.00 overall** — the strongest result across every architecture tested in this project, from the original single-table prototype through every intermediate stage.

```{r criteria-chart, echo=FALSE}
crit <- data.frame(
  criterion = c("Answer correct", "Concise response", "Hallucination free", "Tool correct"),
  pct = c(
    mean(combined$grade_answer, na.rm = TRUE),
    mean(combined$grade_minimal_response, na.rm = TRUE),
    mean(combined$grade_hallucination, na.rm = TRUE),
    mean(combined$grade_tool, na.rm = TRUE)
  )
)
ggplot(crit, aes(x = reorder(criterion, pct), y = pct)) +
  geom_col(fill = "#264653", width = 0.6) +
  geom_text(aes(label = percent(pct, accuracy = 1)), vjust = -0.5, fontface = "bold") +
  coord_flip() +
  scale_y_continuous(limits = c(0, 1.1), labels = percent) +
  labs(title = "Pass rate per grading criterion", x = NULL, y = "Proportion correct") +
  theme_als
```

**99.1% hallucination-free** is the number we'd point to first in a genomics context — almost never does the system state something false with confidence.

```{r category-chart, echo=FALSE, fig.height=7}
cat_scores <- combined %>%
  group_by(category) %>%
  summarise(mean_total = mean(grade_total, na.rm = TRUE), n = n(), .groups = "drop") %>%
  filter(category %in% ALL_CATEGORIES) %>%
  mutate(category = factor(category, levels = ALL_CATEGORIES, labels = ALL_CATEGORY_LABELS))

ggplot(cat_scores, aes(x = reorder(category, mean_total), y = mean_total)) +
  geom_col(fill = "#2A9D8F", width = 0.6) +
  geom_text(aes(label = sprintf("%.2f", mean_total)), hjust = -0.2, fontface = "bold") +
  coord_flip() +
  scale_y_continuous(limits = c(0, 4.3), breaks = 0:4) +
  labs(title = "Average score per category", x = NULL, y = "Average score (out of 4)") +
  theme_als
```

Annotation Trap, Unanswerable, and RVAT — the three categories most directly testing whether the *rvat statistical bridge* and *refusal logic from Stage 3/4* actually work — all sit at a perfect 4.00. Analytical and Complex remain the weakest, around 3.1, and the honest reason is precise multi-step SQL construction — a known limitation, not a mystery.

---

# Live Demo

The working app will be shown live on the SURF HPC server, running the architecture described in Stage 4 above. A few example questions worth trying live, chosen to show the range of what the system can now do that it couldn't at Stage 1:

- *A Stage-1-style single-table question:* "How many high-impact variants are in ABCA4?"
- *A Stage-2-style multi-table question:* "How many female carriers are there in SOD1?"
- *A Stage-3-style routed/refused question:* "What is the allele frequency of VAR_id 30 in Europeans?" *(genuinely unanswerable — watch it refuse correctly rather than guess)*
- *A Stage-4-style real statistics question:* "Run a burden test for SOD1 in ALS cases versus controls."

---

# Where This Goes Next

The clearest, evidence-backed next step is the same pattern that has worked at every stage of this project: when the benchmark shows the model repeatedly struggling with the same *kind* of question, build it a dedicated tool rather than hoping better prompting fixes it. The remaining gap in Analytical and Complex questions is a known, specific, and — based on this project's own track record — solvable problem.

The full codebase, all 111 ground-truth benchmark questions, every benchmark run from this project's history, and this document are all in the project's git repository under `final_app/` and `analysis/`.
