---
title: "Project ALS — Natural Language Database Access"
subtitle: "Benchmarking Local LLMs for Genomic Data Querying"
author: "Projecticum Data Science"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 6
)

library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(knitr)
library(kableExtra)
library(DiagrammeR)

# ── Paths to access final csv data ──────────────────────────
graded_files <- c(
  "~/project_ALS_databse/analysis/benchmark_grading/finalgraded_20260510_141103.csv",
  "~/project_ALS_databse/analysis/benchmark_grading/finalgraded_20260511_010604.csv",
  "~/project_ALS_databse/analysis/benchmark_grading/finalgraded_20260511_093437.csv"
)
# ────────────────────────────────────────────────────────────

combined <- do.call(rbind, lapply(graded_files, read.csv))

averaged <- combined %>%
  group_by(model, id, category) %>%
  summarise(
    grade_answer           = mean(grade_answer,           na.rm = TRUE),
    grade_minimal_response = mean(grade_minimal_response, na.rm = TRUE),
    grade_hallucination    = mean(grade_hallucination,    na.rm = TRUE),
    grade_sql              = mean(grade_sql,              na.rm = TRUE),
    grade_total            = mean(grade_total,            na.rm = TRUE),
    .groups = "drop"
  )

averaged$id <- factor(averaged$id,
  levels = c("L1","L2","L3","L4","L5",
             "A1","A2","A3","A4","A5",
             "U1","U2","U3","U4","U5"))

model_colors <- c(
  "deepseek-r1:8b" = "#E63946",
  "llama3.1:8b"    = "#F4A261",
  "llama3.2"       = "#2A9D8F",
  "mistral"        = "#457B9D",
  "qwen3:8b"       = "#9B5DE5"
)

category_colors <- c(
  "analytical"   = "#E9C46A",
  "lookup"       = "#2A9D8F",
  "unanswerable" = "#457B9D"
)

criterion_colors <- c(
  "Answer"           = "#E63946",
  "Hallucination"    = "#2A9D8F",
  "Minimal_response" = "#457B9D",
  "SQL"              = "#9B5DE5"
)

theme_als <- theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 15, margin = margin(b = 10)),
    plot.subtitle    = element_text(color = "grey40", size = 11, margin = margin(b = 10)),
    axis.title       = element_text(face = "bold"),
    axis.text        = element_text(color = "grey20"),
    axis.text.x      = element_text(angle = 45, hjust = 1),
    legend.title     = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey90"),
    plot.caption     = element_text(color = "grey50", size = 9, margin = margin(t = 10))
  )
```

---

# 1. Project Background

## What is Project MinE?

[Project MinE](https://projectmine.com) is an international collaboration coordinated by the research group of **Kevin Kenna** at UMC Utrecht, involving 27 centers across 21 countries. The project sequences the genomes of ALS patients to map genetic variants associated with the disease.

The genetic data is stored in **VCF format** inside an SQL database. Accessing this database currently requires:

- Proficiency in SQL querying
- Knowledge of VCF format and genomic annotation
- Familiarity with the R package `rvat`

This creates a significant barrier for researchers who are not bioinformaticians.

## The Goal

Make the database **accessible to a wider group of scientists** using natural language processing — allowing researchers to query the database in plain English, without writing a single line of SQL.

---

# 2. Pipeline Architecture

## Technology Stack

The pipeline is built from three core components:

| Component | Role |
|---|---|
| `rvat` | R package by Kenna Lab — opens and manages the `.gdb` genomic database |
| `querychat` | R package by Posit — converts natural language → SQL queries |
| `rollama` / `ollama` | Runs a local LLM to power the NLP |

## How It Works

The user types a plain English question. The local LLM interprets the intent and generates an SQL query via `querychat`. The query is executed against the database and the results are returned as a natural language answer.

```{r pipeline-diagram, echo=FALSE, fig.cap="Current querychat pipeline"}
grViz("
digraph current_pipeline {
  graph [layout = dot, rankdir = LR, fontname = 'Helvetica']
  node [fontname = 'Helvetica', fontsize = 12, style = filled, shape = roundedbox]
  
  User    [label = 'Researcher\\n(plain English)', fillcolor = '#E8F4FD', color = '#457B9D']
  LLM     [label = 'Ollama\\n(local LLM)',         fillcolor = '#E1D5E7', color = '#9B5DE5']
  QC      [label = 'querychat\\n(NL → SQL)',        fillcolor = '#FFF2CC', color = '#E9C46A']
  DB      [label = 'GDB Database\\n(SQLite)',        fillcolor = '#FFE6CC', color = '#F4A261']
  Answer  [label = 'Answer\\nto researcher',         fillcolor = '#D5E8D4', color = '#2A9D8F']

  User   -> LLM    [label = 'question']
  LLM    -> QC     [label = 'generates SQL']
  QC     -> DB     [label = 'executes query']
  DB     -> LLM    [label = 'results']
  LLM    -> Answer [label = 'summarises']
}
")
```

## Example Query

```r
library(ellmer)
library(querychat)
library(rvat)
library(DBI)

# Connect to the ALS genomic database
gdb <- gdb(rvat_example("rvatData.gdb"))

# Set up the chat client with a local LLM
ollama_client <- chat_ollama(
  model  = "llama3.1:8b",
  params = ellmer::params(temperature = 0.1, num_predict = 200)
)

# Initialise QueryChat
qc <- QueryChat$new(
  data_source      = gdb,
  table_name       = "varInfo_synthetic",
  client           = ollama_client,
  data_description = data_description,
  extra_instructions = extra_instructions
)

# Launch interactive console
qc$console()
```

---

# 3. The Database

## Structure

The `varInfo_synthetic` table used for benchmarking contains **1802 variants** with the following key columns:

```{r db-structure, echo=FALSE}
col_info <- data.frame(
  Column = c("VAR_id", "CHROM", "POS", "gene_name", "HighImpact",
             "ModerateImpact", "Synonymous", "CADDphred",
             "PolyPhen", "SIFT", "ALS_1–ALS_5", "Control_1–Control_5"),
  Description = c(
    "Unique variant identifier",
    "Chromosome (e.g. chr1)",
    "Genomic position",
    "Gene the variant falls in",
    "1 = high impact (stop-gain, frameshift); 0 otherwise",
    "1 = moderate impact (missense); 0 otherwise",
    "1 = synonymous (silent); 0 otherwise",
    "Deleteriousness score; higher = more deleterious; missing = '.'",
    "D=probably damaging, P=possibly damaging, B=benign; missing = '.'",
    "D=deleterious, T=tolerated; missing = '.'",
    "Genotype of 5 ALS patients (0=hom ref, 1=het, 2=hom alt)",
    "Genotype of 5 controls (same encoding)"
  )
)

kable(col_info, align = c("l", "l")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) %>%
  row_spec(0, bold = TRUE, background = "#457B9D", color = "white")
```

## Important Data Rules

> - A variant **cannot** be both `Synonymous` and `HighImpact` simultaneously
> - Missing values in `CADDphred`, `PolyPhen`, and `SIFT` are stored as `'.'` — **not** NULL
> - `PolyPhen` and `SIFT` are **computational predictions only** — not evidence of pathogenicity
> - Genotype values indicate zygosity only — **not** disease association

---

# 4. Benchmark Design

## Question Categories

The benchmark was designed to test three types of queries:

```{r benchmark-table, echo=FALSE}
bm <- data.frame(
  Category = c("Lookup (L1–L5)", "Analytical (A1–A5)", "Unanswerable (U1–U5)"),
  Description = c(
    "Simple filtering and counting queries — e.g. 'How many variants are in NEK1?'",
    "Aggregations and calculations — e.g. 'What is the total allele burden in cases vs controls?'",
    "Questions the database cannot answer — tests hallucination avoidance"
  ),
  Example = c(
    "How many variants in TARDBP are predicted deleterious by SIFT?",
    "What is the average allele frequency for synonymous, moderate, and high-impact variants separately?",
    "What is the average age of ALS cases?"
  )
)

kable(bm, align = c("l", "l", "l")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) %>%
  row_spec(0, bold = TRUE, background = "#2A9D8F", color = "white")
```

## Grading Criteria

Each response was graded on **4 criteria** (pass/fail), giving a maximum score of 4 per question:

| Criterion | What it tests |
|---|---|
| `grade_answer` | Is the final answer correct or appropriate? |
| `grade_minimal_response` | Is the response concise (1–3 sentences)? |
| `grade_hallucination` | Is the response free of hallucinations? |
| `grade_sql` | Is the SQL query logically correct? |

## Evaluation Methodology

The benchmark was run **3 times** per model to account for variability. Responses were first **auto-graded** using a local LLM judge (`gemma3`) with the full benchmark document as ground truth, then **manually reviewed** for any rows the auto-grader could not parse.

---

# 5. Models Tested

```{r models-table, echo=FALSE}
models <- data.frame(
  Model = c("llama3.1:8b", "llama3.2", "mistral", "qwen3:8b", "deepseek-r1:8b"),
  Size = c("4.9 GB", "2.0 GB", "4.4 GB", "5.2 GB", "5.2 GB"),
  Notes = c(
    "Preferred by research group; good tool calling",
    "Lighter version of llama3; faster but less capable",
    "Strong instruction following; weaker tool calling",
    "Strong reasoning; good structured output",
    "Reasoning-focused; failed tool calling entirely"
  )
)

kable(models, align = c("l", "l", "l")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) %>%
  row_spec(0, bold = TRUE, background = "#9B5DE5", color = "white")
```

All models were run **locally** via Ollama on the Posit Workbench server — no external API calls, no data leaving the server.

---

# 6. Results

## Overall Performance

```{r plot-overall, echo=FALSE, fig.cap="Overall average score per model (out of 4)"}
model_scores <- averaged %>%
  group_by(model) %>%
  summarise(mean_total = mean(grade_total, na.rm = TRUE), .groups = "drop")

# for usage later in part 7
mistral_score <- model_scores$mean_total[model_scores$model == "mistral"]
llama_score   <- model_scores$mean_total[model_scores$model == "llama3.1:8b"]

ggplot(model_scores, aes(x = reorder(model, mean_total),
                          y = mean_total,
                          fill = model)) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_text(aes(label = round(mean_total, 2)),
            hjust = -0.2, size = 4, fontface = "bold") +
  scale_fill_manual(values = model_colors) +
  scale_y_continuous(limits = c(0, 4.5), breaks = 0:4) +
  coord_flip() +
  labs(
    title    = "Overall average score per model",
    subtitle = "Averaged across 3 benchmark runs, 15 questions each",
    x        = NULL,
    y        = "Average score (out of 4)",
    caption  = "Scoring criteria: answer correctness, response conciseness, hallucination, SQL correctness"
  ) +
  theme_als +
  theme(legend.position = "none", axis.text.x = element_text(angle = 0))
```

## Score by Question Category

```{r plot-category, echo=FALSE, fig.cap="Average score broken down by question category"}
category_scores <- averaged %>%
  group_by(model, category) %>%
  summarise(mean_total = mean(grade_total, na.rm = TRUE), .groups = "drop")

ggplot(category_scores, aes(x = model, y = mean_total, fill = category)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  geom_text(aes(label = round(mean_total, 1)),
            position = position_dodge(width = 0.7),
            vjust = -0.4, size = 3.2, fontface = "bold") +
  scale_fill_manual(
    values = category_colors,
    labels = c("Analytical", "Lookup", "Unanswerable")
  ) +
  scale_y_continuous(limits = c(0, 4.8), breaks = 0:4) +
  labs(
    title    = "Average score per question category per model",
    subtitle = "Averaged across 3 benchmark runs",
    x        = "Model",
    y        = "Average score (out of 4)",
    fill     = "Category",
    caption  = "Lookup = factual queries | Analytical = aggregations | Unanswerable = hallucination tests"
  ) +
  theme_als
```

## Score per Question (Heatmap)

```{r plot-heatmap, echo=FALSE, fig.height=9, fig.cap="Score per question per model — red = low, green = high"}
ggplot(averaged, aes(x = model, y = id, fill = grade_total)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = round(grade_total, 1)),
            size = 3.5, fontface = "bold", color = "grey20") +
  scale_fill_gradient2(
    low      = "#E63946",
    mid      = "#F4A261",
    high     = "#2A9D8F",
    midpoint = 2,
    limits   = c(0, 4),
    name     = "Score\n(0–4)"
  ) +
  scale_x_discrete(position = "top") +
  labs(
    title    = "Score per question per model",
    subtitle = "Averaged across 3 benchmark runs",
    x        = NULL,
    y        = "Question ID",
    caption  = "L = Lookup | A = Analytical | U = Unanswerable"
  ) +
  theme_als +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
        panel.grid  = element_blank())
```

## Score by Grading Criterion

```{r plot-criteria, echo=FALSE, fig.cap="Performance broken down by grading criterion"}
criterion_scores <- averaged %>%
  group_by(model) %>%
  summarise(
    Answer           = mean(grade_answer,           na.rm = TRUE),
    Minimal_response = mean(grade_minimal_response, na.rm = TRUE),
    Hallucination    = mean(grade_hallucination,    na.rm = TRUE),
    SQL              = mean(grade_sql,              na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols      = c(Answer, Minimal_response, Hallucination, SQL),
    names_to  = "criterion",
    values_to = "score"
  )

ggplot(criterion_scores, aes(x = model, y = score, fill = criterion)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  geom_text(aes(label = round(score, 2)),
            position = position_dodge(width = 0.7),
            vjust = -0.4, size = 3, fontface = "bold") +
  scale_fill_manual(
    values = criterion_colors,
    labels = c("Answer correct", "Hallucination free", "Minimal response", "SQL correct")
  ) +
  scale_y_continuous(
    limits = c(0, 1.15),
    breaks = seq(0, 1, 0.25),
    labels = percent
  ) +
  labs(
    title    = "Score per grading criterion per model",
    subtitle = "Averaged across 3 benchmark runs",
    x        = "Model",
    y        = "Proportion correct",
    fill     = "Criterion",
    caption  = "Each criterion scored as pass/fail per question"
  ) +
  theme_als
```

---

# 7. Key Findings

## What Worked Well

- **SQL generation** was generally correct across all models that executed tool calls — the LLMs understood the schema and wrote valid queries
- **Unanswerable questions** were handled better than expected — most models correctly refused questions about missing columns (age, population-specific AF, pathogenicity)
- **llama3.1:8b** showed the lowest hallucination rate, consistent with the research group's prior experience
- The pipeline is **fully local** — no data leaves the server, which is critical for sensitive genomic data

## What Needs Improvement

- **Response verbosity** — all models struggled with conciseness, often returning long rambling responses instead of a direct 1–3 sentence answer
- **Instruction following** — smaller local models consistently ignored formatting rules despite explicit instructions in the system prompt
- **Mistral's tool calling** — Mistral scored highest overall but does not actually execute queries; it writes SQL as text without running it, making it unsuitable for this pipeline despite good scores
- **Ambiguous questions** — models rarely asked for clarification when terms like "most important" were used

## The Mistral Paradox

> Mistral achieved the highest benchmark score 
> (`r round(mistral_score, 2)`/4) but is the least suitable model for production use, as it does not call the querychat tool and therefore never actually queries the database. 
> **llama3.1:8b**, preferred by the research group, scores lower 
> (`r round(llama_score, 2)`/4) but correctly executes the full pipeline.

This highlights an important limitation of automated benchmarking — **scores do not always reflect real-world usefulness**.

---

# 8. Limitations

- The benchmark was run on a **synthetic dataset** — specific values differ from the real Project MinE database, so only SQL logic correctness (not exact values) was fully verifiable
- Only **5 ALS patients and 5 controls** were included in the synthetic data — real analyses involve thousands of samples
- The auto-grader (`gemma3`) introduced some inconsistency — approximately 10–15% of rows required manual review
- All models were tested with identical prompts — further prompt engineering per model might improve individual performance

---

# 9. Future Work — MCP Agent

## Why querychat Has Limitations

The current `querychat` pipeline is limited to **one input table** at a time. Real genomic analyses often require:

- Joining variant data with phenotype data (`pheno` table)
- Running RVAT association tests
- Querying multiple tables simultaneously

## The Proposed Solution: MCP Agent

The next phase of the project involves building a **Model Context Protocol (MCP) agent** in Python that can:

- Expose **multiple data sources** as separate tools the LLM can call
- Integrate **RVAT functions** as callable tools (not just SQL)
- Handle **multi-step reasoning** across tables
- Give full control over response formatting

```{r mcp-diagram, echo=FALSE}
grViz("
digraph mcp_agent {
  graph [layout = dot, rankdir = TB, fontname = 'Helvetica']
  node [fontname = 'Helvetica', fontsize = 12, style = filled, shape = roundedbox]

  User   [label = 'User question',             fillcolor = '#E8F4FD', color = '#457B9D']
  LLM    [label = 'Ollama (Llama 3.1)',         fillcolor = '#E1D5E7', color = '#9B5DE5']
  MCP    [label = 'MCP Agent (Python)',         fillcolor = '#FFF2CC', color = '#E9C46A']
  SQL    [label = 'Tool: VCF / SQL database',   fillcolor = '#FFE6CC', color = '#F4A261']
  RVAT   [label = 'Tool: RVAT rare variant analysis', fillcolor = '#FFE6CC', color = '#F4A261']
  Pheno  [label = 'Tool: Phenotype / metadata', fillcolor = '#FFE6CC', color = '#F4A261']
  Answer [label = 'Answer to researcher',       fillcolor = '#D5E8D4', color = '#2A9D8F']

  User -> LLM  [label = 'question']
  LLM  -> MCP  [label = 'decides tool calls']
  MCP  -> SQL
  MCP  -> RVAT
  MCP  -> Pheno
  SQL  -> LLM  [label = 'results']
  RVAT -> LLM
  Pheno -> LLM
  LLM  -> Answer [label = 'synthesises']

  { rank = same; SQL; RVAT; Pheno }
}
")
```

## Why MCP Over querychat?

| | querychat | MCP Agent |
|---|---|---|
| Multiple data sources | ✗ | ✓ |
| RVAT integration | ✗ | ✓ |
| Response formatting control | Limited | Full |
| Extensibility | Limited | High |
| Setup complexity | Low | Medium |

## MCP first impression

> schrijf ff wat shit op over python en moeilijkheidsgraad

## MCP findings

> zo veel mogelijk shit dat je al weet en dan uitleggen erover

---

---
  
# 9b. Alternative Approaches — Extending querychat
  
Before committing to the full MCP agent architecture, two lighter-weight 
alternatives are worth exploring. Both stay entirely within R, avoiding the 
added complexity of a Python backend and an HTTP layer between Shiny and the 
agent.

## Approach 1 — Forking querychat

### How querychat works internally

querychat is relatively thin. At its core it does three things:
  
1. Builds a system prompt describing the target tables schema and any extra 
   instructions
2. Passes the users question + that system prompt to an `ellmer` chat client
3. Executes the returned SQL against the database and returns the result

The key entry point is `QueryChat$new()`, which accepts a single `table_name` 
and builds its schema description from that one table. The multi-table 
limitation is therefore not architectural — it is simply a consequence of how 
the system prompt is constructed.

### The workaround

By forking the querychat source and modifying the schema-building step, it 
should be possible to:
  
- Accept a **list of tables** (e.g. `varInfo`, `pheno`) instead of a single 
`table_name`
- Describe **all table schemas and their relationships** in the system prompt 
(primary keys, how tables join)
- Allow the LLM to generate **multi-table SQL** (JOIN queries) rather than 
single-table queries

The rest of the pipeline — the ellmer chat client, the SQL execution step, the 
Shiny integration — would remain unchanged.

### What needs investigating

> - Where exactly in the querychat source is the system prompt constructed? 
  >   Which function or R6 method handles schema introspection?
  > - Does querychat validate or sanitise the returned SQL before executing it? 
  >   Would JOIN queries pass through unmodified?
  > - Are there any hardcoded assumptions about a single table elsewhere in the 
>   codebase (e.g. in how results are returned or displayed)?
  > - How does querychat handle execution errors — would a malformed JOIN cause 
>   a silent failure or a visible error?
  
### Relevant source to look at
  
```r
# Install the development version to inspect source
# install.packages("pak")
pak::pak("posit-dev/querychat")

# Browse the R6 class definition
querychat:::QueryChat
```

---
  
## Approach 2 — Using ellmer directly
  
### What this means
  
Since querychat is itself a wrapper around `ellmer`, it is possible to bypass 
querychat entirely and use ellmers tool-calling interface directly. This gives 
full control over:

- How many tables are described in the system prompt
- What tools the LLM can call (SQL queries, RVAT functions, metadata lookups)
- How responses are formatted before being returned to Shiny

The tradeoff is that you are rebuilding what querychat already provides — SQL 
execution, result handling, conversation history — from scratch in R.

### Sketch of the approach

```r
library(ellmer)
library(DBI)
library(rvat)

# Connect to database
gdb <- gdb(rvat_example("rvatData.gdb"))

# Define a SQL execution tool the LLM can call
run_sql <- function(query) {
  tryCatch(
    dbGetQuery(gdb@db, query),
    error = function(e) paste("SQL error:", e$message)
  )
}

# Define an RVAT tool the LLM can call
run_rvat_burden <- function(gene, group_col = "ALS_status") {
  # placeholder — actual rvat call goes here
  paste("RVAT burden test for", gene, "not yet implemented")
}

# Build a system prompt describing all available tables and tools
system_prompt <- "
You are a genomic data assistant with access to two tools:
  - run_sql: executes a SQL query against the variant database
  - run_rvat_burden: runs a rare variant burden test for a given gene

The database contains the following tables:

varInfo columns: VAR_id, CHROM, POS, gene_name, HighImpact, ModerateImpact,
Synonymous, CADDphred, PolyPhen, SIFT, ALS_1 through ALS_5, Control_1 
through Control_5.

pheno columns: sample_id, ALS_status, age_at_onset, site_of_onset, sex.

varInfo and pheno share no direct join key in SQL — use RVAT for 
case/control analyses that require both tables.

Always respond in 1-3 sentences. Do not hallucinate columns that are not listed.
"

# Initialise ellmer chat with tool definitions
client <- chat_ollama(
  model  = "llama3.1:8b",
  system = system_prompt,
  params = ellmer::params(temperature = 0.1)
)

# Register tools
client$register_tool(tool(
  run_sql,
  "Execute a SQL query against the genomic variant database.",
  query = type_string("A valid SQLite query.")
))

client$register_tool(tool(
  run_rvat_burden,
  "Run a rare variant burden test for a given gene using RVAT.",
  gene     = type_string("The gene name to test."),
  group_col = type_string("Column to use for case/control grouping.")
))

# Ask a question
client$chat("How many high-impact variants are in NEK1?")
```

### What needs investigating

> - Does `ellmer` support registering multiple tools and does `llama3.1:8b` 
>   reliably select between them? Tool selection failures were already observed 
>   with some models in the benchmark.
> - What is the correct ellmer syntax for `register_tool` and `tool()` in the 
>   current version — the API has been evolving and documentation is sparse.
> - How does conversation history work in ellmer when tool calls are involved — 
>   does it automatically maintain context across multi-turn queries?
> - Can the Shiny integration from the current querychat app be adapted to work 
>   with a raw ellmer client, or does it depend on querychat-specific internals?

---

## Comparison of approaches

| | Fork querychat | ellmer directly | MCP Agent |
|---|---|---|---|
| Multi-table SQL | ✓ (with modification) | ✓ | ✓ |
| RVAT integration | ✗ (SQL only) | ✓ | ✓ |
| Stays in R | ✓ | ✓ | ✗ |
| Shiny compatibility | ✓ (minimal changes) | Needs testing | Needs HTTP layer |
| Implementation effort | Low–medium | Medium | High |
| Modifies upstream package | Yes | No | No |

The fork approach is the quickest path to multi-table SQL. The ellmer approach 
is more powerful and avoids dependency on querychats internals, but requires 
more groundwork. Both are worth prototyping before deciding whether the full 
MCP architecture is necessary.

---

---

# 9c. Implementation — MultiTableSource

## Approach

Rather than forking the querychat package or rebuilding the pipeline in Python,
the multi-table limitation was resolved by implementing querychat's existing
`DataSource` interface directly. This required no changes to the querychat
source code and no new dependencies — everything stays within R.

The entry point was `normalize_data_source()`, which contains an early exit:

```r
if (is_data_source(data_source)) {
    return(data_source)
}
```

Where `is_data_source()` simply checks:

```r
inherits(x, "DataSource")
```

This means any R6 class inheriting from `querychat:::DataSource` is accepted
as-is, bypassing all single-table wrapping logic. The interface required is
five methods: `get_schema()`, `execute_query()`, `test_query()`, `get_data()`,
and `get_db_type()`.

## What Was Built

`MultiTableSource` is a ~60 line R6 class that:

- Accepts a `DBIConnection` and a **vector** of table names instead of a single
  table
- Overrides `get_schema()` to call querychat's internal `get_schema_impl()`
  once per table and concatenate the results — this is what gets injected into
  the LLM system prompt
- Treats the first table as the primary display table for `get_data()` and
  `test_query(require_all_columns = TRUE)`
- Delegates `execute_query()` and `test_query()` directly to DBI

One additional change was required: the default querychat prompt template
contains the line `"For security reasons, you may only query this specific
table."` This is hardcoded outside any conditional block and would cause the
LLM to ignore the second table despite it being present in the schema. The
fix was to copy the template, replace that line with `"You may query any of
the tables described in the schema above."`, and pass the custom template path
via the `prompt_template` argument to `QueryChat$new()`.

Two minor errors were encountered during setup:

- `gdb@db` is not a valid slot — the `gdb` object itself is the
  `DBIConnection` and should be passed directly
- `QueryChat$new()` attempts to build an id string via
  `sprintf("querychat_%s", table_name)`, which fails when `table_name` is a
  vector; resolved by passing `id = "querychat_multi"` explicitly

## Results

The implementation was tested against three questions:

| Question | Table queried | Result |
|---|---|---|
| What is the average age of ALS cases? | `pheno` | 60.1 years — correct |
| What is the sex distribution of the cohort? | `pheno` | 51.6% / 48.4% — correct |
| How many high-impact variants are in SOD1? | `varInfo_synthetic` | 4 — incorrect |

The LLM correctly routed each query to the appropriate table without any
explicit routing logic — schema description alone was sufficient. All
previously unanswerable benchmark questions involving sample metadata (age,
sex, population) are now answerable.

## What This Means for the Benchmark

Several questions graded as `unanswerable` in the original benchmark — and
correctly refused by the models — are now answerable with this implementation.
The benchmark results therefore represent a lower bound on what the pipeline
can achieve; a re-run with `MultiTableSource` would be needed to assess
performance on pheno-level questions.

## Limitations

- The `pheno` table has a column also named `pheno` (1 = ALS case, 0 =
  control), which is potentially ambiguous for the LLM. This should be
  clarified in `data_description`.
- Dashboard filtering (`querychat_update_dashboard`) still operates on the
  primary table only (`varInfo_synthetic`) — filtered views of the `pheno`
  table are not supported in the current implementation.
- The two tables are not directly joinable via a shared column value. `IID`
  values in `pheno` (e.g. `ALS1`) correspond to column *names* in
  `varInfo_synthetic` (e.g. `ALS_1`), not column values — cross-table
  analyses therefore cannot be expressed as a single SQL JOIN.
  
---

# 10. Shiny Application

## What is Shiny?

[Shiny](https://shiny.posit.co/) is an R framework for building interactive web applications directly from R code — no HTML, CSS, or JavaScript required. In the context of this project, Shiny provides the **graphical user interface** that researchers interact with, sitting on top of the querychat pipeline.

Instead of typing queries in an R console, users get a clean web interface with a chat sidebar and a live data table — making the tool accessible to non-technical researchers.

## Basic Shiny Structure

The minimal Shiny app wrapping our QueryChat pipeline looks like this:

```{r shiny-example, eval=FALSE}
library(shiny)
library(bslib)
library(querychat)

# gooi hier een basic pipeline van shiny
```

## Findings

> shiny info, problemen waar je tegen bent gelopen.

Belangrijke puntjes voor presentatie:

- layout problemen?
- UI 

## Potential UI Improvements

> reset knopje, makkelijk berijkbaar chatbox, etc

---

# 11. Proposed Production Architecture

Combining the MCP agent backend with the Shiny frontend, the full proposed production pipeline would look as follows:
  
```{r mcp-full-diagram, echo=FALSE, fig.cap="Proposed production architecture"}

grViz("
digraph MCP_pipeline {

  graph [layout = dot, rankdir = TB, fontname = 'Helvetica']
  
  node [fontname = 'Helvetica', fontsize = 12, style = filled, shape = roundedbox]

  # User layer
  User [label = 'Researcher\\n(plain English question)', 
        fillcolor = '#E8F4FD', color = '#457B9D']

  # Frontend layer
  Shiny [label = 'Shiny Web App\\n(R frontend)', 
         fillcolor = '#D5E8D4', color = '#2A9D8F']

  # Agent layer
  MCP [label = 'MCP Agent\\n(Python orchestrator)', 
       fillcolor = '#FFF2CC', color = '#E9C46A']

  # LLM layer
  LLM [label = 'Ollama — Llama 3.1\\n(local LLM)', 
       fillcolor = '#E1D5E7', color = '#9B5DE5']

  # Tool layer
  SQL  [label = 'Tool: VCF / SQL DB\\nvariant queries', 
        fillcolor = '#FFE6CC', color = '#F4A261']
  RVAT [label = 'Tool: RVAT (R)\\nrare variant analysis', 
        fillcolor = '#FFE6CC', color = '#F4A261']
  META [label = 'Tool: Phenotype / metadata\\ncohort information', 
        fillcolor = '#FFE6CC', color = '#F4A261']

  # Result layer
  Synth [label = 'Result synthesis\\n(Llama 3.1)', 
         fillcolor = '#E1D5E7', color = '#9B5DE5']

  # Output layer
  Answer [label = 'Answer + data table\\ndisplayed in Shiny', 
          fillcolor = '#D5E8D4', color = '#2A9D8F']

  # Edges — question flow
  User   -> Shiny  [label = 'types question', color = '#457B9D']
  Shiny  -> MCP    [label = 'HTTP request', color = '#2A9D8F']
  MCP    -> LLM    [label = 'interprets intent', color = '#9B5DE5']
  LLM    -> MCP    [label = 'decides tool calls', color = '#9B5DE5']
  MCP    -> SQL    [color = '#F4A261']
  MCP    -> RVAT   [color = '#F4A261']
  MCP    -> META   [color = '#F4A261']
  SQL    -> Synth  [color = '#F4A261']
  RVAT   -> Synth  [color = '#F4A261']
  META   -> Synth  [color = '#F4A261']
  Synth  -> Shiny  [label = 'synthesized answer', color = '#2A9D8F']
  Shiny  -> Answer [label = 'renders response', color = '#2A9D8F']
  Answer -> User   [color = '#457B9D']

  # Rank groupings
  { rank = same; SQL; RVAT; META }
}
")
```

## Why This Architecture?

```{r architecture-table, echo=FALSE}
arch <- data.frame(
  Layer = c("Shiny (R)", "MCP Agent (Python)", "Ollama / Llama 3.1", "Tools"),
  Role = c(
    "User-facing interface — chat sidebar, results table, SQL display",
    "Orchestrates tool calls, manages context, routes requests",
    "Understands natural language, decides which tools to call",
    "Execute queries, run RVAT analyses, retrieve metadata"
  ),
  `Why this choice` = c(
    "Native R integration, familiar to the research group, querychat-compatible",
    "MCP is purpose-built for multi-tool LLM orchestration",
    "Local model — no data leaves the server, free to run",
    "Separating tools means new data sources can be added without rebuilding the pipeline"
  )
)

kable(arch, col.names = c("Layer", "Role", "Why this choice"),
      align = c("l", "l", "l")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) %>%
  row_spec(0, bold = TRUE, background = "#9B5DE5", color = "white")
```

---

# 12. Conclusions

The proof-of-concept demonstrates that local LLMs can successfully translate natural language queries into SQL for genomic variant databases. The pipeline is:

- **Functional** — queries execute correctly for most question types
- **Private** — fully local, no external API calls
- **Accessible** — researchers without SQL knowledge can retrieve data

However, the current implementation has clear limitations that motivate the move to an MCP agent architecture for the production version.

**Recommended model for current pipeline: `llama3.1:8b`** — lowest hallucination rate, correct tool calling, consistent with research group preference.

---

*This document was generated as part of the Projecticum Data Science course. All genomic data used in benchmarking is synthetic.*
